|

### 2024-08-20 reproducibility/manuscript/manuscript.qmd
docs(manuscript): update text
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

>>>>

Next, we evaluated the ability of the Pyro&thinsp;-Velocity models to recover
bi-fate or multi-fate trajectories by considering clones belonging to monocyte
and neutrophil lineages or using the entire dataset. Not surprisingly, both
methods showed their limitations, given that they are not designed to model and
capture multiple fate trajectories ([@suppfig-lineage-tracing-2]c-d). The
recovered vector fields and cell latent (shared) time show inconsistent
directions for both methods. However, Pyro&thinsp;-Velocity identified higher
uncertainty for some of the problematic regions. Furthermore, we discovered that
vector field and shared time estimation could be significantly improved by
extending the original kinetic model by allowing each gene a time lag for the
transcriptional activation and a baseline expression level potentially different
than 0 (Pyro&thinsp;-Velocity Model 2, [@sec-methods])
([@fig-lineage-tracing]c,d, [@suppfig-graphical-models]b,c). This extended model
recapitulated a vector field consistent with the clonal progression vector field
for most branches with some spurious predictions within the intermediate cell
state between stem cells and Megakaryocytes and most terminal Neutrophils
(bi-fate cosine similarity: $0.62$, multi-fate cosine similarity: $0.63$).
However, the proposed uncertainty estimation accurately flagged these
subpopulations to warn users of potential false discoveries. Notably, this
extended model significantly improved the correlation of cell shared time with
the clonal fate potency (bi-fate $\rho=0.65$, multi-fate $\rho=0.74$) and also
improved the correlation with Cytotrace fate potency (bi-fate $\rho=0.931$,
multi-fate $\rho=0.614$). Remarkably, the shared time per cell from
Pyro&thinsp;-Velocity also outperformed Cytotrace for developmental order
prediction ([@suppfig-lineage-tracing]a,c). Importantly, this extended model
retained its ability to accurately identify cell lineage trajectories and shared
time for all the other datasets analyzed above ([@suppfig-model-2-result]).
Taken together, our approach using LARRY barcoded datasets established the first
quantitative benchmark of RNA velocity-based cell fate trajectory inference and
provides an exciting new framework for continued optimization of multi-fate
predictions in the future.
